Notes

### 2022-09-01 chapters/abstract.tex
Añade contenido en Ch.5; alguna tabla

\noindent This work is framed within the field of Functional Data Analysis, a branch of statistics in which the objects of interest are random functions in a functional space instead of, say, random points in \(\R^p\). Due to the infinite-dimensional nature of the data, the most common \(L^2\)-models for functional linear and logistic regression present some complications that require further simplifications, usually in the form of regularization or dimensionality reduction.

We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of reproducing kernel Hilbert spaces (RKHS's). These recently-proposed models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest, which has a tractable parametric form. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be estimated through Markov chain Monte Carlo methods, which do not require a complete specification of the posterior density.

We derive several prediction strategies from the approximate posterior distribution, including a Bayesian-motivated variable selection procedure. We show through a comprehensive set of experiments that these methods are competitive against other usual alternatives in terms of predictive performance, both in simulated examples and real data sets. Overall, our proposed model is simple with regard to interpretation and implementation, and has an added flexibility thanks to the ambient Bayesian framework.\\

>>>>

\noindent This work is framed within the field of Functional Data Analysis, a branch of statistics in which the objects of interest are random functions in a functional space, instead of, say, random points in \(\R^p\). Due to the infinite-dimensional nature of the data, the most common \(L^2\)-models for functional linear and logistic regression present some complications that require further simplifications, usually in the form of regularization or dimensionality reduction.

In this thesis we propose a novel Bayesian approach for functional linear and logistic regression models, based on the theory of reproducing kernel Hilbert spaces (RKHS's). These recently-proposed models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest, which has a tractable parametric form. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be estimated through Markov chain Monte Carlo methods, which do not require a complete specification of the posterior density.

We derive several prediction strategies from the approximate posterior distribution, including a Bayesian-motivated variable selection procedure. We show through a comprehensive set of experiments that these methods are competitive against other usual alternatives in terms of predictive performance, both in simulated examples and real data sets. Overall, our proposed model is simple with regard to interpretation and feasible with regard to implementation, while enjoying the added flexibility of an ambient Bayesian framework.\\

### 2022-09-01 chapters/appendix-code.tex
Añade contenido en Ch.5; alguna tabla

\begin{outcomment}
  Aquí irá un pequeño resumen de la estructura del código y los archivos, y un mini-tutorial de cómo usarlo para reproducir los experimentos. No debería ocupar más de una página.

  \begin{itemize}
    \item  Comentar cómo usar el código, las versiones de los paquetes, y los notebooks.
    \item Contar que son clases a la sklearn y compatibles con ska-fda. Enlazar al github.
    \item Se usan 4 cores.
    \item El proceso es reproducible (semilla 2022).
  \end{itemize}
\end{outcomment}

>>>>

\begin{outcomment}
  Aquí irá un pequeño resumen de la estructura del código y los archivos, y un mini-tutorial de cómo usarlo para reproducir los experimentos.

  \begin{itemize}
    \item Características de la máquina donde se ejecuta. Se usan 4 cores.
    \item  Comentar cómo usar el código, las versiones de los paquetes, y los notebooks (decir que hay más gráficas y análisis y cosas así).
    \item Contar que son clases a la sklearn y compatibles con ska-fda. Enlazar al github.
    \item El proceso es reproducible (semilla 2022).
  \end{itemize}
\end{outcomment}

### 2022-09-01 chapters/background-related-work.tex
Añade contenido en Ch.5; alguna tabla

When \(x\in\R\), we can use Monte Carlo algorithms such as \textit{rejection sampling}. In a nutshell, this method samples uniformly from the area under \(f(x)\) by means of an auxiliary density that encompasses this area. Specifically, one needs to find a density function \(g\) with \(\supp f \subset \supp g\) and \(f \leq Mg\) for some \(M>1\). Then, sampling from \(g\) and accepting the proposal, say \(x\), with probability \(f(x)/Mg(x)\) yields an approximate set of samples from \(f\). Nevertheless, these kinds of algorithms are rapidly affected by the so-called \textit{curse of dimensionality} (see Figure \ref{fig:curse_dimensionality}), so new techniques are needed when sampling from multidimensional distributions. This is where Markov chain Monte Carlo (MCMC) methods come into play.

>>>>

When \(x\in\R\), we can use Monte Carlo algorithms such as \textit{rejection sampling}. In a nutshell, this method samples uniformly from the area under \(f(x)\) by means of an auxiliary density that encompasses this area. Specifically, one needs to find a density function \(g\) with \(\supp f \subset \supp g\) and \(f \leq Mg\) for some \(M>1\). Then, sampling from \(g\) and accepting the proposal, say \(x\), with probability \(f(x)/Mg(x)\) yields an approximate set of samples from \(f\). Nevertheless, these kinds of algorithms are rapidly affected by the so-called \textit{curse of dimensionality} (see the schematic in Figure~\ref{fig:curse_dimensionality}), so new techniques are needed when sampling from multidimensional distributions. This is where Markov chain Monte Carlo (MCMC) methods come into play.

### 2022-09-01 chapters/conclusions.tex
Añade contenido en Ch.5; alguna tabla

\begin{outcomment}
  Algo que no he comentado en ninguna parte es que el tiempo de ejecución de nuestros modelos es bastante alto (del orden de 30-40 segundos por ejecución) frente al de los algoritmos de comparación (unos pocos segundos en general). Como es algo negativo, no sé muy bien cómo ponerlo, si comentarlo de pasada aquí en las conclusiones, o dedicarle un poco más de espacio en el capítulo de experimentos.\\
\end{outcomment}

\begin{outcomment}
  Algunos comentarios generales:
  \begin{itemize}
    \item Destacar las ventajas de lo bayesiano: flexibilidad, incorporación de información con distintas priors, etc.
    \item Destacar la interpretabilidad y sencillez de nuestro modelo. En general, conseguimos resultados competitivos usando modelos con baja dimensión.
    \item Los resultados en regresión logística parecen en media un poco mejores que en regresión lineal.
    \item No se recomienda usar la media como estadístico de resumen de la distribución a posteriori.
    \item La búsqueda de hiperparámetros en CV no ha sido demasiado grande. Probablemente con más recursos computacionales se podrían conseguir mejores resultados.
  \end{itemize}
\end{outcomment}

>>>>

\begin{outcomment}
  Algunos comentarios generales:
  \begin{itemize}
    \item Destacar las ventajas de lo bayesiano: flexibilidad, incorporación de información con distintas priors, etc. Computationally feasible model.
    \item Destacar la interpretabilidad y sencillez de nuestro modelo. En general, conseguimos resultados competitivos usando modelos con baja dimensión. Aunque el modelo final es simple, la teoría subyacente no lo es. Modelo simple útil para practitioners. También se presta a post-interpretación (ej. si salen dos modas)
    \item Algo del tiempo de ejecución. Se podría aproximar mejor el error medio con más repeticiones aleatorias (en lugar de 10), pero sería muy costoso computacionalmente.
    \item Los resultados en regresión logística parecen en media un poco mejores que en regresión lineal.
    \item No se recomienda usar la media como estadístico de resumen de la distribución a posteriori. En general, los que usan la posteriori entera son más estables.
    \item La búsqueda de hiperparámetros en CV no ha sido demasiado grande. Probablemente con más recursos computacionales se podrían conseguir mejores resultados.
  \end{itemize}
\end{outcomment}

### 2022-09-01 chapters/experiments.tex
Añade contenido en Ch.5; alguna tabla

First and foremost, we consider \(T=\{\text{mean, median, mode}\}\) for the summarize-then-predict approach to prediction. In this way in linear regression we have 4 prediction methods (one for each statistic and one for the predict-then-summarize approach, which will we call ``\textit{posterior\_mean}'' for short) and 3 variable selection methods (one for each statistic), while in logistic regression there are 5 prediction methods and 3 variable selection procedures. In each case, after variable selection is performed, we use a penalized multiple linear/logistic regression method to generate the corresponding predictions. All in all, we are looking at 7 (8 in the case of logistic regression) prediction methods, and although all of them are derived from a single MCMC run, we will treat them as separate for the purposes of experimentation. In what follows we relax the notation and use the expressions ``model'', ``prediction method'' and ``algorithm'' almost interchangeably when referring to these 7 (or 8) cases.

@@

We consider a set of functional regressors common to linear and logistic regression problems. They are 4 Gaussian processes (GP's), each with a different kernel function:
\begin{enumerate}
  \item A Brownian motion \(K_1(t,s)=\min\{t,s\}\).
  \item A fractional Brownian motion \(K_2(t,s)=1/2(s^{2H} + t^{2H} - |t-s|^{2H})\), with Hurst parameter \(H=0.8\).
  \item An Ornstein-Uhlenbeck process \(K_3(t,s)=e^{-|t-s|}\).
  \item A Gaussian kernel \(K_4(t,s)=e^{-(t-s)^2/2\nu^2}\), with \(\nu=0.2\).
\end{enumerate}
Also, when applicable, we fix a variance \(\sigma^2=0.5\) for the error terms \(\epsilon\).

@@

\paragraph{Logistic regression data sets.} Again we consider three different types of data sets,  with a common value of \(\alpha_0=-0.5\).
\begin{itemize}
  \item Four logistic RKHS models with the same functional parameter as in the linear regression case (one for each GP). Specifically,
  \[
  \P(Y=1\mid X) = \frac{1}{e^{0.5 +5X(0.1) - X(0.4) - 10X(0.8)}}.
  \]
  \item Four logistic \(L^2\)-models with the same coefficient function as in the linear regression case.
  \item A ``mixture'' model based on non-GP regressors, in which we mix regressors from two different GP's and label them according to their origin. Firstly, we consider a homoscedastic case to distinguish between a standard Brownian motion and a Brownian motion with a mean function that is zero until \(t=0.5\), and then becomes \(m(t)=0.75t\). Secondly, we consider a heteroscedastic case to distinguish between a standard Brownian motion and a Brownian motion with variance \(2\), that is, with kernel \(K(t,s)=2\min\{t,s\}\).
\end{itemize}
In this case there are again three real data sets. The first one is a subset of the Medflies data set \citep{carey1998relationship}, consisting on samples of the number of eggs laid daily by 534 flies over 30 days, trying to predict if their longevity is high or low. The second one is the Berkeley Growth Study data set \citep{tuddenham1954physical}, which records the height of 54 girls and 39 boys over 31 different points in their lives. Finally, we have selected a subset of the Phoneme data set \citep{hastie1995penalized}, based on 200 digitized speech frames over 128 equispaced points to predict the phonemes ``aa'' and ``ao''.

@@

We have adopted a visual approach to presenting the experimentation results, using colored graphs to help visualize them. Alternatively, tables representing these results can be consulted instead in Appendix~\ref{ch:tables}.

We show the mean and standard deviation of the RMSE obtained across the 10 random runs, depicting our models in orange and the comparison algorithms in blue. We also show the global mean of all the comparison algorithms with a dashed vertical line, and we exclude extreme negative results from this mean to avoid distortion. Moreover, we separate complete prediction algorithms from two-stage methods, the latter being the ones that perform variable selection or dimension reduction prior to a multiple linear/logistic regression method.

We do not expect our models to defeat all the comparison algorithms in all cases (and indeed they do not), but we intend to show that they are competitive methods that on average match or improve other usual alternatives. After all, \textit{there is no such thing as a free lunch}\footnote{The phrase alludes to the No Free Lunch theorem, which loosely speaking asserts that all machine learning methods are equivalent when their performance is averaged across all possible problems. However, the practical implications of this statement are questionable \citep[see e.g.][]{giraud2005toward}.}!

\section{Functional linear regression}

@@

Note that our Bayesian models have a higher standard deviation, partly because there is an intrinsic randomness in the methods (apart from the train/test splits), and it can be the cause of the occasional worse performance. In relation to this, we observe that the methods that use the mean as a summary statistic tend to perform much worse. This is because the mean is very sensitive to outliers, so if at some point there is a chain that randomly deviates from the rest, the mean of the posterior distribution will be greatly impacted. In particular, very high values of the parameter \(\sigma^2\) were reported sometimes, producing misleading results when averaged with the rest of the values.

@@

\section{Functional logistic regression}

@@

Continuing with Figure~\ref{fig:clf_emcee_l2}, we see that in the \(L^2\) case the results are again promising, since our models score consistently on or above the mean of the reference models, and in many cases surpassing most of them. Moreover, the overall accuracy of all methods is low (below 60\%), so this is indeed a difficult problem in which even small increases in accuracy are relevant.

@@

As for the real data sets, in Figure~\ref{fig:clf_emcee_real} we see positive results in general, obtaining in most cases accuracies well above the mean of the reference models, and sometimes above most of them. However, our variable selection methods seem to be a bit worse in this case. Again, the models that use \textit{emcee\_mean} are the exception, and in all these data sets they perform steadily worse than the rest of our models.

@@

In this section we present some experiments that are not directly related to measuring the prediction performance relative to other methods.

@@

One requirement that our model should satisfy is that it be able to recover the true parameter function when the underlying model is a RKHS model. This is generally true when the value of \(p\) in our model and the true value of \(p\) coincide, but what happens when we change the value of \(p\) in the model?

Take for example a data set generated according to the formula \(Y=-5X(0.1) + 10X(0.8) + \epsilon\), with \(\epsilon\sim \mathcal N(0, 0.5)\). Figure~\ref{fig:beta_trace_3} shows the resulting posterior distribution of the parameters \(b=(\beta_1, \beta_2, \beta_3)\) and \(\tau=(\tau_1, \tau_2, \tau_3)\) for a model with 3 components.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.8\textwidth]{beta_trace_3}
  \caption{Posterior distribution in a linear model with \(p=3\).}\label{fig:beta_trace_3}
\end{figure}

As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function (based for example in the mode of the posterior distributions) will not be perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.

In contrast, if we consider now a model with \(p=4\) with the same data, we might obtain posterior distributions such as the ones in Figure~\ref{fig:beta_trace_4}. In this situation two coefficients should go to zero, but that is no longer the case. For example, while the green component has a high density around 0, it also has a considerable mass around 10, effectively ``competing'' with the red component. This is a manifestation of the label switching issue, caused in this case by an excessive number of degrees of freedom in the model.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.8\textwidth]{beta_trace_4}
  \caption{Posterior distribution in a linear model with \(p=4\).}\label{fig:beta_trace_4}
\end{figure}

There is still another possible situation, one in which there is no label switching but the four components are non-negligible in the estimated function. This can happen because the components exploit the additional freedom and ``work together'', so to speak. In this way we could obtain an estimate that might not resemble the true coefficient function, but that could have a very low prediction error. However, this could also work to our detriment and cause the estimated function to be worse prediction-wise than simpler alternatives. This phenomenon is expected to strengthen as the difference between the true and assumed value of \(p\) grows larger.

\incomment{Me gustaría añadir aquí un último experimento en el que se muestran gráficas de evolución del RMSE o del Accuracy según va variando alguno de los hiperparámetros (especialmente el valor de p y el número de ejemplos de entrenamiento).}

@@

The first thing we tried was to make the parameter \(p\) part of the Bayesian model, imposing a categorical prior distribution on it. As explained in Chapter~\ref{ch:model-choice} this approach led to many complications and generally poor results. However, the idea of letting the model select how many components it should have based on the data is an interesting one, and in some situations this could greatly improve the results of variable selection procedures based on other model selection criteria.

We also tried changing some prior distributions. For instance, we swapped the improper prior distribution on \(\sigma^2\) for a Cauchy distribution on \(\R^+\). The results were similar, but we ultimately discarded this approach because it introduced yet another hyperparameter into the model with no clear benefit. In addition, we tried to impose a \(\beta\) distribution on each of the time instants \(t_j\), selecting the hyperparameters in a semi-automatic way so that most of the density fell in the areas where the variance of the process was higher (and thus the effect of individual trajectories was more noticeable). This approach does work, but in our estimation the computational overhead introduced by the beta density function (about 1.5x increase in execution time) is not worth the mild improvements, if any.

>>>>

First and foremost, we consider \(T=\{\text{mean, median, mode}\}\) for the summarize-then-predict approach to prediction. In this way in linear regression we have 4 prediction methods (one for each statistic and one for the predict-then-summarize approach, which will we call ``\textit{posterior\_mean}'' for short) and 3 variable selection methods (one for each statistic), while in logistic regression there are similarly 5 prediction methods and 3 variable selection procedures. In each case, after variable selection is performed, we use a penalized multiple linear/logistic regression method to generate the corresponding predictions. All in all, we are looking at 7 (8 in the case of logistic regression) prediction methods, and although all of them are derived from a single MCMC run, we will treat them as separate for the purposes of the experimentation. In what follows we relax the notation and use the expressions ``model'', ``prediction method'' and ``algorithm'' almost interchangeably when referring to these 7 (or 8) cases.

@@

We consider a set of functional regressors common to linear and logistic regression problems. They are 4 Gaussian processes (GP's), each with a different kernel function:
\begin{enumerate}
  \item BM.\hspace{0.3em} A Brownian motion \(K_1(t,s)=\min\{t,s\}\).
  \item fBM.\hspace{0.3em} A fractional Brownian motion \(K_2(t,s)=1/2(s^{2H} + t^{2H} - |t-s|^{2H})\), with Hurst parameter \(H=0.8\).
  \item O-U.\hspace{0.3em} An Ornstein-Uhlenbeck process \(K_3(t,s)=e^{-|t-s|}\).
  \item Gaussian.\hspace{0.3em} A Gaussian kernel \(K_4(t,s)=e^{-(t-s)^2/2\nu^2}\), with \(\nu=0.2\).
\end{enumerate}
Also, when applicable, we fix a variance \(\sigma^2=0.5\) for the error terms \(\epsilon\).

@@

\paragraph{Logistic regression data sets.} Again we consider three different types of data sets,  with a common value of \(\alpha_0=-0.5\). In this case we randomly permute 10\% of the labels to introduce some noise in the simulations.
\begin{itemize}
  \item Four logistic RKHS models with the same functional parameter as in the linear regression case (one for each GP). Specifically,
  \[
  \P(Y=1\mid X) = \frac{1}{e^{0.5 +5X(0.1) - X(0.4) - 10X(0.8)}}.
  \]
  \item Four logistic \(L^2\)-models with the same coefficient function as in the linear regression case.
  \item A ``mixture'' model based on non-GP regressors, in which we mix regressors from two different GP's and label them according to their origin. Firstly, we consider a homoscedastic case to distinguish between a standard Brownian motion and a Brownian motion with a mean function that is zero until \(t=0.5\), and then becomes \(m(t)=0.75t\). Secondly, we consider a heteroscedastic case to distinguish between a standard Brownian motion and a Brownian motion with variance \(2\), that is, with kernel \(K(t,s)=2\min\{t,s\}\).
\end{itemize}
In this case there are again three real data sets. The first one is a subset of the Medflies data set \citep{carey1998relationship}, consisting on samples of the number of eggs laid daily by 534 flies over 30 days, trying to predict if their longevity is high or low. The second one is the Berkeley Growth Study data set \citep{tuddenham1954physical}, which records the height of 54 girls and 39 boys over 31 different points in their lives. Finally, we have selected a subset of the Phoneme data set \citep{hastie1995penalized}, based on 200 digitized speech frames over 128 equispaced points to predict the phonemes ``aa'' and ``ao''.

@@

We have adopted a visual approach to presenting the experimentation results, using colored graphs to help visualize them. We felt that this was a better way of summarizing a large empirical study such as the one we have carried out. Alternatively, tables representing these results can be consulted in Appendix~\ref{ch:tables}.

We show the mean and standard deviation of the score obtained in each case across the 10 random runs, depicting our models in orange and the comparison algorithms in blue. We also show the global mean of all the comparison algorithms with a dashed vertical line, and we exclude extreme negative results from this mean to avoid distortion. Moreover, we separate complete prediction algorithms from two-stage methods, the latter being the ones that perform variable selection or dimension reduction prior to a multiple linear/logistic regression method.

We do not expect our models to defeat all the comparison algorithms in all cases (and indeed they do not), but we intend to show that they are competitive methods that on average match or improve the performance of other usual alternatives. After all, \textit{there is no such thing as a free lunch}\footnote{The phrase alludes to the No Free Lunch theorem, which loosely speaking asserts that all machine learning methods are equivalent when their performance is averaged across all possible problems. However, the practical implications of this statement are questionable \citep[see e.g.][]{giraud2005toward}.}!

\section{Functional linear regression}\label{sec:experiments-linear}

@@

Note that some of our Bayesian models have a higher standard deviation, partly because there is an intrinsic randomness in the methods (apart from the train/test splits), and it can be the cause of the occasional worse performance. In relation to this, we observe that the methods that use the mean as a summary statistic tend to perform much worse. This is because the mean is very sensitive to outliers, so if at some point there is a chain that randomly deviates from the rest, the mean of the posterior distribution will be greatly impacted. In particular, very high values of the parameter \(\sigma^2\) were reported sometimes, producing misleading results when averaged with the rest of the values.

@@

\section{Functional logistic regression}\label{sec:experiments-logistic}

@@

Continuing with Figure~\ref{fig:clf_emcee_l2}, we see that in the \(L^2\) case the results are again promising, since our models score consistently on or above the mean of the reference models, and in many cases surpassing most of them. The predict-then-summarize approaches (\textit{emcee\_posterior\_mean} and \textit{emcee\_posterior\_vote}) are particularly good in this case, and in general have low standard errors. Moreover, the overall accuracy of all methods is low (below 60\%), so this is indeed a difficult problem in which even small increases in accuracy are relevant.

@@

As for the real data sets, in Figure~\ref{fig:clf_emcee_real} we see positive results in general, obtaining in most cases accuracies well above the mean of the reference models, and sometimes above most of them. In particular, the predict-then-summarize tend to have a good performance and achieve a lower standard error across executions, which is a trend that we also saw in the simulated data sets. However, our variable selection methods seem to be a bit worse in this case. Again, the models that use \textit{emcee\_mean} are the exception, and in all these data sets they perform steadily worse than the rest of our Bayesian models.

@@

In this section we present some experiments that are not directly related to measuring the prediction performance of our models relative to other methods.

@@

One requirement that our model should satisfy is that it be able to recover the true parameter function when the underlying model is a RKHS model. This is generally the case when the value of \(p\) in our model and the true value of \(p\) coincide, but what happens when we change the value of \(p\) in the model?

Take for example a data set generated according to the formula \(Y=-5X(0.1) + 10X(0.8) + \epsilon\), with \(\epsilon\sim \mathcal N(0, 0.5)\). Figure~\ref{fig:beta_trace_3} shows the resulting posterior distribution of the parameters \(b=(\beta_1, \beta_2, \beta_3)\) and \(\tau=(\tau_1, \tau_2, \tau_3)\) for a model with 3 components. As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function (based for example in the mode of the posterior distributions) will not be perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.8\textwidth]{beta_trace_3}
  \caption{Estimated posterior distribution for \(\beta\) and \(\tau\) in a linear model with \(p=3\) (left) and the corresponding trace evolution for 700 iterations in the MCMC sampler (right).}\label{fig:beta_trace_3}
\end{figure}

In contrast, if we consider now a model with \(p=4\) with the same data, we might obtain posterior distributions like the ones in Figure~\ref{fig:beta_trace_4}. In this situation two coefficients should go to zero, but that is no longer the case. For example, while the green component has a high density around 0, it also has a considerable mass around 10, effectively ``competing'' with the red component. This is a manifestation of the label switching issue, caused in this case by an excessive number of degrees of freedom in the model.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=.8\textwidth]{beta_trace_4}
  \caption{Estimated posterior distribution for \(\beta\) and \(\tau\) in a linear model with \(p=4\) (left) and the corresponding trace evolution for 700 iterations in the MCMC sampler (right).}\label{fig:beta_trace_4}
\end{figure}

There is still another possible situation, one in which there is no label switching but the estimated function has four non-negligible components. This can happen because the different components exploit the additional freedom and ``work together'', so to speak. In this way we might obtain an estimate that does not resemble the true coefficient function, but that has a very low prediction error. However, this could also work to our detriment and cause the estimated function to be worse prediction-wise than simpler alternatives. This phenomenon is expected to strengthen as the difference between the true and assumed value of \(p\) grows larger.

\subsection*{Dependence on \(p\)}

Another thing we wanted to look at was the dependence of the final prediction result on the chosen value of \(p\), especially when there is no concept of ``components'' in the underlying model. We can take for example the homoscedastic mixture data set in the logistic regression problem, and fix the parameter \(\eta=0.01\). Then the corresponding mean accuracies (in 10 repetitions) for RKHS models with \(p=1,2,\dots 10\) components are shown in Figure~\ref{fig:dependence_acc_p}, along with their standard errors (which are arguably not very informative).

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{mixture_dependence_acc_p}
  \caption{Mean accuracy in 10 repetitions for our logistic RKHS methods as a function of \(p\), using \(\eta=0.01\) and the homoscedastic mixture data set. The corresponding standard errors are shown in faded colors.}\label{fig:dependence_acc_p}
\end{figure}

It would appear that in the methods that use the whole posterior distribution are more stable and somewhat independent of the value of \(p>1\). The other methods show a slight downward trend as \(p\) increases (although not so much in the variable selection methods), and in general their best results are obtained at \(p=2\) and \(p=8\). We expect that this effect or some small variation of it will remain valid in other situations; however, a profound study of this would be the subject of a different experiment altogether.

\subsection*{Execution times}

We summarize below the mean execution time for each independent run of our CV experiment, grouping the simulated data sets by the underlying data generation strategy (since the results within these groups were very similar).

\begin{table}[ht!]
  \centering
  \begin{tabular}{lcccccc}
\toprule
  &            RKHS &           \(L^2\) &           GBM &        Tecator & Moisture & Sugar \\
\midrule
Linear regression & 260 & 245 & 264 & 230 & 264 & 285\\
\bottomrule

  &            RKHS &           \(L^2\) &           Mixture &        Medflies & Growth & Phoneme \\
\toprule
Logistic regression & 277 & 263 & 246 & 250 & 218 & 277\\
\bottomrule
\end{tabular}
  \caption{Mean execution times (in minutes) for the independent runs of the CV experiments in Sections~\ref{sec:experiments-linear} and~\ref{sec:experiments-logistic}, grouped when possible by the underlying data generation strategy.}
\end{table}

The reference algorithms used were all fast methods that can complete their whole CV loop in just a few seconds, so any comparison with them in this regard is uninteresting. However, this outcome was to be expected, since in general the MCMC methods are known to have a high computational cost in exchange for the simplicity in modeling they provide. An individual MCMC run is not that expensive time-wise, but even for a modest 5-fold cross-validation loop the effects add up quickly to a considerable amount of time. In Figure~\ref{fig:dependence_time_n_train} we show the evolution of the execution time versus the number of training samples in a generic case, which not surprisingly has a more or less linear behavior.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.6\textwidth]{l2_dependence_time_n_train}
  \caption{Mean training times of a logistic RKHS model, averaged across 10 independent runs, as a function of the number of training samples. The values are surrounded by the corresponding standard deviations. }\label{fig:dependence_time_n_train}
\end{figure}

@@

The first thing we tried was to make the parameter \(p\) part of the Bayesian model, imposing a categorical prior distribution on it. As explained in Chapter~\ref{ch:model-choice}, this approach led to many complications and generally poor results. However, the idea of letting the model select how many components it should have based on the data is an interesting one, and in some situations this could greatly improve the results of variable selection procedures based on other model selection criteria.

We also tried changing some prior distributions. For instance, we swapped the improper prior distribution on \(\sigma^2\) for a Cauchy distribution on \(\R^+\). The results were similar, but we ultimately discarded this approach because it introduced yet another hyperparameter into the model with no clear benefit. In addition, we tried to impose a \(\beta\) distribution on each of the time instants \(t_j\), selecting the hyperparameters in a semi-automatic way so that most of the density fell in the areas where the variance of the process was higher (and thus the effect of individual trajectories was more noticeable). This approach does work, but in our estimation the computational overhead introduced by the beta density function (sometimes resulting in a 1.5x increase in execution time) is not worth the mild improvements, if any.

### 2022-09-01 chapters/introduction.tex
Añade contenido en Ch.5; alguna tabla

\begin{enumerate}[1.]
  \item To perform a brief but thorough literature review that contextualizes functional linear and logistic regression in statistics and machine learning, exploring the predominant models and techniques.
  \item To propose a novel RKHS-based functional model for linear and logistic regression that builds on existing work and focuses on simplicity, both in terms of interpretation and implementation.
  \item To describe and implement a general Bayesian approach for parameter estimation within the suggested model.
  \item To introduce the tools needed to specify and implement the proposed model, mainly reproducing kernel Hilbert spaces and Markov chain Monte Carlo methods.
  \item To carry out a comprehensive experimental study to test these new models and compare them to existing methods, both in simulations and in real-world scenarios.
\end{enumerate}

It is beyond the scope of this thesis to provide an extensive review of FDA as a whole, or to delve too deeply into the details and inner workings of most models and techniques mentioned. Nevertheless, references are provided frequently throughout the text to point the interested reader towards more specialized resources.

@@

In Chapter~\ref{ch:background} we summarize the relevant literature related to our problem, along with a review of the basics of RKHS's and MCMC methods. Chapter~\ref{ch:bayesian} is devoted to explaining the Bayesian methodology and the functional regression models we propose. Then, in Chapter~\ref{ch:model-choice} we present a short discussion of theoretical and computational details that have led to the concrete specification of the model, as well as some validation techniques. The empirical results of the experimentation are contained in Chapter~\ref{ch:experiments}. Lastly, the conclusions drawn from this work and future paths of research are presented in Chapter~\ref{ch:conclusions}.

>>>>

\begin{enumerate}[1.]
  \item To perform a brief but thorough literature review that contextualizes functional linear and logistic regression within statistics and machine learning, exploring the predominant models and techniques.
  \item To propose a novel RKHS-based functional model for linear and logistic regression that builds on existing work and focuses on simplicity, both in terms of interpretation and implementation.
  \item To describe and implement a general Bayesian approach for parameter estimation within the suggested model.
  \item To introduce the tools needed to specify the proposed model and put it into practice, mainly reproducing kernel Hilbert spaces and Markov chain Monte Carlo methods.
  \item To carry out an extensive experimental study to test these new models and compare them to existing methods, both in simulations and in real-world scenarios.
\end{enumerate}

It is beyond the scope of this thesis to provide a complete review of FDA as a whole, or to delve too deeply into the details and inner workings of most models and techniques mentioned. Nevertheless, references are provided frequently throughout the text to point the interested reader towards more specialized resources.

@@

In Chapter~\ref{ch:background} we summarize the relevant literature related to our problem, along with a review of the basics of RKHS's and MCMC methods. Chapter~\ref{ch:bayesian} is devoted to explaining the Bayesian methodology and the functional regression models we propose. Then, in Chapter~\ref{ch:model-choice} we present a short discussion of theoretical and computational details that have led to the concrete specification of the model, as well as some validation techniques. The empirical results of the experimentation are contained in Chapter~\ref{ch:experiments}. Lastly, the conclusions drawn from this work and future paths of research are reviewed in Chapter~\ref{ch:conclusions}.
